RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2015-01-02 manuscripts/manuscript.Rmd
First part of data extension section.

## Data annotation with custom metadata extensions

@@

## RDF manipulation and SPARQL queries

The most powerful way to take advantage of the machine-readable nature
of the metadata involves the construction of SPARQL queries. This
approach can exploit the reasoning and logical deduction capacity of
formal vocabularies (ontologies). For a proper introduction to SPARQL
queries, see @W3C_2014 or @Hartig_2012).

The real expressiveness of SPARQL is better illustrated through more
complex examples such as the one shown in the supplementary materials
(which can also be run from R using the command: `demo("sparql", "RNeXML")`).
This demonstration constructs a cladogram based on the taxonomic ranks
provided in the NeXML. SPARQL allows us to make explicit use of ontological
concepts such as manipulating sub-classes (i.e. that "Genus" is a subclass 
of "Family"), or joining multiple predicates (i.e. retrieve the NCBI identifier
of the OTU that has taxon rank of "Order").

## Extending the NeXML standard through metadata annotation.

No data standard can anticipate all the uses and extensions researchers
may one day need. Researchers have repeatedly adapted the existing NEXUS
file format to accommodate encoding new kinds of information, but in the
process break compatibility between different flavors of NEXUS files, with
few reliable methods to distinguish between these flavors.  NeXML provides
a solution to this problem through the use of metadata annotation.

Here we illustrate this process using the example of stochastic character
mapping [@Huelsenbeck_2003]. A stochastic character map is simply
an annotation of the branches on a phylogeny, assigning each section
of each branch to a particular "state" (typically of a morphological
characteristic).

@Bollback_2006 provides a widely used stand-alone software implementation
of this method in the software `simmap`, which modified the standard
Newick tree format to express this additional information. This can
break compatibility with other software [^3], and creates a format that
cannot be interpreted without additional information describing this
convention.  By contrast, the NeXML extension is not only backwards
compatible but contains a precise and machine-readable description of
what it is encoding.

In this example, we illustrate how the additional information required
to define a stochastic character mapping (a `simmap` mapping) in NeXML.

>>>>

## Data annotation with custom metadata

@@

Because NeXML expresses all metadata using the RDF standard, and
stores them compliant with RDFa, they can be extracted as an RDF graph
using off-the-shelf tools, and the RDF graph obtained in this way can
be queried in SPARQL (see @W3C_2014 or @Hartig_2012), the query
language for RDF graphs, using general purpose triple stores and
SPARQL query engines. The `RNeXML` vignette illustrates this with
examples, and the package also comes with a demonstration that can be
run from R using the following command: `demo("sparql", "RNeXML")`).

## Using metadata to extend the NeXML standard

No data standard can anticipate all the uses and extensions
researchers may one day need. As the various "flavors" of the NEXUS
and other data formats show, this often results in extensions to
widely used exchange formats that almost invariably break
interoperability between such extensions and tools that were designed
to read one but not another. NeXML was designed to prevent this by
means of a highly flexible metadata system, while not sacrificing 
strict validation of syntax and structure.

Here we illustrate how NeXML's metadata system, and `RNeXML`'s API to
it, can be used to record and share a type of phylogenetic data not
taken into account when NeXML was designed. Specifically, we use the
example of stochastic character maps [@Huelsenbeck_2003], which assign
sections of each branch in a phylogeny to a particular "state"
(typically of a morphological characteristic). The current de-facto
format for sharing stochastic character maps is a (non-interoperable)
modification of the standard Newick tree format created by a software
tool widely used for creating such maps, `simmap` [@Bollback_2006].
